A general and efficient representation of Ancestral Recombination Graphs

### 2023-09-15 paper.tex
Rewrite gARG

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an overly
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

\section*{Genome ARGs}

We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations. A ``genome'' in our usage corresponds to the
haploid genetic material present in a particular individual or cell.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

A gARG is then a graph in which nodes represent
haploid genomes, edges represent
genetic inheritance between an ancestor and a descendant, and
annotations specify coordinates of the genomic intervals
over which genetic inheritance occurs.
% FIXME WE're missing the important distinction here that the pedigree
% has one node per ind, whereas a gARG has two nodes.
We can therefore view a gARG for a given set of individuals
as a subset of their pedigree.
An example of this for diploids
(so each individual is represented by two nodes)
is illustrated in Fig.~\ref{fig-arg-in-pedigree}.
Each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \textsf{a} and \textsf{b},
inherited from parents $I_3$ and $I_4$. Genome \textsf{a} is the product of
recombining $I_3$'s two genomes \textsf{e} and \textsf{f} at position 2,
whereas \textsf{b} was inherited directly from $I_4$'s genome \textsf{g} without
recombination.

@@

The topology of a gARG specifies that genetic inheritance occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination (and any other process
that results in the partial inheritance of a descendant genome
from an ancestor) we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
Continuing the example of genome \textsf{a} in Fig.~\ref{fig-arg-in-pedigree},
the recombination at position 2 means that it inherits the (half-closed) genome
interval $[0, 2)$ from \textsf{e} and it inherits $[2, 10)$ from \textsf{f};
these intervals are shown attached to the corresponding graph edges.

% important to show that this is a formal model and the concrete definition
% will help some readers?
% FIXME - the oriented forest definition is wrong, we need to be more
% concrete about the fact that nodes are integers.
Formally, we define a gARG as a sets of nodes $N$ and edges $E$.
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
which genome $c$ inherits from $p$. The most fundamental operation for
an ARG data structure is the process of recovering the corresponding sequence
of local trees along the genome. Concretely, we can describe a local tree
at position $x$ as an oriented forest $\pi = ( \pi_1, \hdots, \pi_{|N|} )$,
where the entry $\pi_j$ is the index
of the parent of $j$ and $\pi_j = \emptyset$ if $j$ is not present in the
tree~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
We use a forest to describe the local ancestry at a given
position so that we allow for cases where not all leaves have
found a most recent common ancestor, resulting in multiple roots.
To construct $\pi$ at position $x$, set
$\pi_i = \emptyset$ for $i \in \{1, \hdots, |N|\}$.
% FIXME - this is wrong
Then, traversing over all nodes of the tree,
at each node
$u$, find an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$.
At the end of this process, $\pi$ fully describes the local tree.

There are some interesting consequences to these definitions. Firstly,
there is
no limitation on mating system, ploidy, or age structure. By making the
basic unit of analysis an individual's \emph{genomes} (i.e.\ two per individuals
for diploids), any pattern of inheritance among mixed ploidy can be
accommodated.
Secondly, because we record the intervals of inheritance
between genomes (and not causal \emph{events}),
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations (e.g.\ Figure~\ref{fig-arg-in-pedigree}), and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, ``local'' trees that we generate for a particular position on the
genome are not explicitly concerned with ``coalescence'' (nodes at which
one or more samples find a common ancestor); for example, node \noderef{e}
is present but not coalescent in the left-most tree of Figure~\ref{fig-arg-in-pedigree} (see the ``Precision of recombination information'' section for more
discussion). Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g. node \noderef{a}
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

[FIXME find somewhere natural for this. The last clause is too vague - what
are the implications?]
Alternative gARG encodings split transmitted intervals and associate each with
a new node, resulting in a single genome being represented by a group of
multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
efficiency of storage.

In this section we have considered the nodes in a gARG as separate from events
in a generative process. Nevertheless, gARGs are commonly constructed to
include only those genomes associated with specific events. For example,
\cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''. In an eARG however, events are even
more explicitly embedded in the data structure.

@@

The amount of information about recombination that is inferrable
is the subject of much research
has a long history [summarise this in a couple of sentences?]
\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
(equivalent to the ``bricked tree sequence'' of
\cite{nowbandegani2023extremely}), where we consider the unique sets of
samples that coalesce at a node in the ARG over a particular
genome interval, and the fundamental limitations they place
on ARG inference.

>>>>

The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[MORE][]{mathieson2020ancestry}, including
varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
However, this broad usage means there is disagreement on the
exact object being defined. This causes problems, for example,
not just when evaluating inferred ARGs, but even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer}.
Perhaps most importantly, the direct linkage to
the original stochastic process definition leads to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.

@@

\section*{Genome ARGs}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
In this usage a ``genome'' is equivalent to the definition of a
``haplotype'' used by~\cite{shipilina2023origin}, which we avoid because
it already has many interpretations, and in particular carries the implication of
specific mutational patterns [CITES] that persist over
many generations.
% A ``genome'' in our usage corresponds to the
% haploid genetic material present in a particular individual or cell.
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.

A genome ARG (gARG) is a graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
% This has two purposes: clarify that we're not talking about Shipilina
% et als edges, and also get "Graph theory" in here
(Note that we are using the term ``edge'' in its standard
graph-theoretic sense
and not the specialised interpretation of
\cite{shipilina2023origin}.)
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
% TODO we should change the notation for inheritance intervals here,
% as we use I to index individuals.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see section XXX for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals $[\ell, r)$ over
which genome $c$ inherits from $p$. Thus, each topological connection between
 a parent and child in the graph is ``annotated'' with set of
inheritance intervals $I$.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
In this diploid example, each genome is inherited from one
of the individual's parents,
and may be the recombined product of that parent's two genomes.
For example, individual $I_1$ in Fig.~\ref{fig-arg-in-pedigree}
has two genomes \noderef{a} and \noderef{b},
inherited from parents $I_3$ and $I_4$.
Genome \noderef{b} was inherited directly from $I_4$'s genome \noderef{g} without
recombination, whereas
\noderef{a} is the product of
recombining $I_3$'s two genomes \noderef{e} and \noderef{f} at position 2.
This recombination at means \noderef{a} inherits the (half-closed) genome
interval $[0, 2)$ from \noderef{e} and it inherits $[2, 10)$ from \noderef{f};
these intervals are shown attached to the corresponding graph edges.

% The most fundamental operation for
% an ARG data structure is to recover the embedded local tree
% at some position $x$ along the genome.
The relationship between an ARG and the corresponding local trees
is somewhat subtle (see section XXX), and it useful to have
a concrete definition and process for reconstruction.
We can describe a local tree as an
oriented tree~\cite[p.\ 461]{knuth11combinatorial},
a sequence of integers $\pi_1\pi_2\dots$ where $\pi_u$ denotes the parent of
node $u$~\citep{kelleher2013coalescent,kelleher2014coalescent,kelleher2016efficient}.
If $u$ is a root
% I'm not sure if this is a helpful clarification, and whether it'll just
% confuse people. It depends on some tricky details.
(or not reachable from the samples)
in the local tree at position $x$, $\pi_u = -1$.
Technically, this structure is a ``forest'', as it allows us to have
multiple disconnected trees, which occurs when all samples have not
found a most recent common ancestor. Oriented trees also allow nodes
to have  any number of children from one (see section XXX for more
discussion on such ``unary nodes'' in the local trees), to arbitrarily
large polytomies.
To recover the local tree $\pi$ at position $x$  then, we begin by
setting $\pi_u = -1$ for each $u\in N$. Then, for each sample
node in $S$ we trace its path rootwards through the
ARG for position $x$, and record this path in $\pi$.
Specifically, at a given node $u$,
we an edge $(c, p, I) \in E$ such that $u = c$ and $x \in I$, and set
$\pi_c \leftarrow p$. The set $u \leftarrow p$, and repeat
until either $\pi_u \neq -1$ (indicating we have traversed this section
of the ARG already on the path from another sample) or there
is no matching outbound edge (indicating we are at a root).

There are some important consequences to these definitions.
Firstly, there is
no limitation on mating system, ploidy, or age structure in a gARG.
By making the
basic unit of analysis an individual's genomes,
any pattern of inheritance among mixed ploidy can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance such as
gene conversions,
multiple simultaneous recombinations
(e.g.\ Figure~\ref{fig-arg-in-pedigree}),
and so on can be directly expressed.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity, including ``unary''
nodes with a single child (see Section XXX).
% are not explicitly concerned with ``coalescence'' (nodes at which
% one or more samples find a common ancestor); for example, node \noderef{e}
% is present but not coalescent in the left-most tree of
% Figure~\ref{fig-arg-in-pedigree}
% (see section XXX for more discussion).
Fourth, recombination is modelled in terms of
its outcomes in terms of inheritance intervals among parental genomes;
thus we will have a child node (the recombinant; e.g. node \noderef{a}
Figure~\ref{fig-arg-in-pedigree}) and two (or more; see section XXX)
paternal genomes from which this child has inherited segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the evolutionary processes of genetic inheritance and
recombination, and is agnostic to the details of those processes.
% The paragraph above is a bit weak and muddled, needs a few more passes.
% In particular, the "Finally" bit doesn't really seem to follow
% from the others. Is there a more direct way we can address these points ?

% In this section we have considered the nodes in a gARG as separate from events
% in a generative process. Nevertheless, gARGs are commonly constructed to
% include only those genomes associated with specific events. For example,
% \cite{mathieson2020ancestry} define an ARG as ``a subset of the pedigree...
% contain[ing] only those edges along which inherited segments of DNA have been
% transmitted, and only those nodes corresponding to ancestors in which there was
% a recombination or coalescence event.''. In an eARG however, events are even
% more explicitly embedded in the data structure.

@@

The amount of information about recombination that is inferrable
is the subject of much research
has a long history [summarise this in a couple of sentences?]
\cite{shipilina2023origin} discuss the idea of a ``haplotype block''
(equivalent to the ``bricked tree sequence'' of
\cite{nowbandegani2023extremely},
% TODO look this up. I expect it's basically the same??
and the block-based ARG data structure of
\citep{palamara2016argon}),
here we consider the unique sets of
samples that coalesce at a node in the ARG over a particular
genome interval, and the fundamental limitations they place
on ARG inference.

% [FIXME find somewhere natural for this. The last clause is too vague - what
% are the implications?]
% Alternative gARG encodings split transmitted intervals and associate each with
% a new node, resulting in a single genome being represented by a group of
% multiple nodes (or ``blocks'' \citep{palamara2016argon}), with implications for
% efficiency of storage.
